The Leaky Semicolon

### 2021-11-01 paper/abstract.tex
accepted!

To support sequential composition while targeting modern hardware, we enrich
the standard event-based approach with \emph{preconditions} and
\emph{families of predicate transformers}.
%
% When calculating the precondition of an event $e$ in $\sem{S_1;S_2}$, from
% $\sem{S_2}$ is derived by applying a predicate transformer chosen based on
% events of $\sem{S_1}$ upon which $e$ depends.
When calculating $\sem{S_1;S_2}$, the predicate transformer applied to the
precondition of an event $e$ from $\sem{S_2}$ is chosen based on the set of
events in $\sem{S_1}$ upon which $e$ depends.  We apply this approach to two
existing memory models.
% : ``Modular Relaxed Dependencies'' for C11 and
% ``Pomsets with Preconditions.''

>>>>

To support sequential composition while targeting modern hardware, we enrich
the standard event-based approach with \emph{preconditions} and
\emph{families of predicate transformers}.
%
% When calculating the precondition of an event $e$ in $\sem{S_1;S_2}$, from
% $\sem{S_2}$ is derived by applying a predicate transformer chosen based on
% events of $\sem{S_1}$ upon which $e$ depends.
When calculating the meaning of $(S_1;S_2)$, the predicate transformer applied to the
precondition of an event $e$ from ${S_2}$ is chosen based on the set of
events in ${S_1}$ upon which $e$ depends.  We apply this approach to two
existing memory models.
% : ``Modular Relaxed Dependencies'' for C11 and
% ``Pomsets with Preconditions.''

### 2021-11-01 paper/drf.tex
removing todos. Integrated consistency and causal strengtheining.
accepted!

% \section{Local Data Race Freedom and Sequential Consistency}
\section{LDRF-SC for \PwTmcaTITLE{}}
\label{sec:sc}

\todo{Remove this section?}

\begin{changed}
  In this appendix, we establish a \drfsc{} for \PwTmca{2}.  We prove an
  \emph{external} result, where the notion of \emph{data-race} is independent
  of the semantics itself.  Since every \PwTmca{2} is also a \PwTmca{1}, the
  result also applies there.  Our result is also \emph{local}.  Using
  \citeauthor{Dolan:2018:BDR:3192366.3192421}'s
  [\citeyear{Dolan:2018:BDR:3192366.3192421}] notion of \emph{Local Data Race
    Freedom (LDRF)}.

@@

\bookmark{Definitions}
\begin{changed}
  \paragraph{Program Order}
  Let $\sempomca{2}{}$ be defined by applying the construction of
  \reflem{lem:po} to $\semmca{2}{}$.  We consider only \emph{complete}
  pomsets.  For these, we derive program order on compound events as follows.
  By \reflem{lem:compound:phantom:exists}, if there is a compound event
  $\aEv$, then there is a phantom event $\cEv\in\fmrginv{\aEv}$ such that
  $\labelingForm(\cEv)$ is a tautology.  If there is exactly one tautology,
  we identify $\aEv$ with $\cEv$ in program order.  If there is more than one
  tautology, \reflem{lem:sequential:simple}, below, shows that it suffices to
  pick an arbitrary one---we identify $\aEv$ with the $\cEv\in\fmrginv{\aEv}$
  that is minimal in program order.
\end{changed}
For example, consider \jmm{} causality test case 2, with an added write to $z$:
\begin{gather*}
  \tag{$\ddagger\ddagger$}\label{eq6}
  \begin{gathered}
    \PR{x}{r}\SEMI
    \PW{z}{1}\SEMI
    \PR{x}{s}\SEMI
    \IF{r{=}s}\THEN \PW{y}{1}\FI
    \PAR
    \PW{x}{y}
    \\
    \hbox{\begin{tikzinline}[node distance=1em and 1.5em]
        \phevent{p1}{\DR{x}{1}}{}
        \event{z}{\DW{z}{1}}{right=of p1}
        \phevent{p2}{\DR{x}{0}}{right=of z}
        \event{a3}{\DW{y}{1}}{right=of p2}
        \event{b1}{\DR{y}{1}}{right=3em of a3}
        \event{b2}{\DW{x}{1}}{right=of b1}
        \rf{a3}{b1}
        \po{b1}{b2}
        \pox[out=-15,in=-170]{p1}{z}
        \pox[out=-15,in=-170]{z}{p2}
        \pox[out=-15,in=-170]{p2}{a3}
        \pox[out=-15,in=-170]{b1}{b2}
        \event{a1}{\DR{x}{1}}{above=of z}
        \mrg{p1}[pos=.3]{a1}
        \mrg{p2}[right]{a1}      
        \rf{b2}{a1}
        \pox{a1}{z}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}
% For TC2, we have: 
% \begin{gather*}
%   \PR{x}{r}\SEMI
%   \PR{x}{s}\SEMI
%   \IF{r{=}s}\THEN \PW{y}{1}\FI
%   \PAR
%   \PW{x}{y}
%   \\
%   \hbox{\begin{tikzinline}[node distance=1.5em]
%       \event{a1}{\DR{x}{1}}{}
%       % \event{a2}{\DR{x}{1}}{right=of a1}
%       \event{a3}{\DW{y}{1}}{right=of a1}
%       % \po{a1}{a3}
%       % \po[out=-20,in=-160]{a1}{a3}
%       \event{b1}{\DR{y}{1}}{right=3em of a3}
%       \event{b2}{\DW{x}{1}}{right=of b1}
%       \rf{a3}{b1}
%       \po{b1}{b2}
%       % \rf[out=169,in=11]{b2}{a2}
%       \rf[out=169,in=11]{b2}{a1}
%       \phevent{p2}{\DR{x}{1}}{below=of a1}
%       \phevent{p1}{\DR{x}{1}}{left=of p2}
%       \mrg{p1}[pos=.3]{a1}
%       \mrg{p2}[right]{a1}      
%       \pox[out=-15,in=-170]{b1}[below]{b2}
%       \pox[out=-15,in=-170]{p1}[below]{p2}
%       \pox{p2}[below]{a3}
%     \end{tikzinline}}
%   \\
%   \hbox{\begin{tikzinline}[node distance=1.5em]
%       \event{a1}{\DR{x}{0}}{}
%       % \event{a2}{\DR{x}{1}}{right=of a1}
%       \event{a3}{\DW{y}{1}}{right=of a1}
%       % \po{a1}{a3}
%       % \po[out=-20,in=-160]{a1}{a3}
%       \event{b1}{\DR{y}{1}}{right=3em of a3}
%       \event{b2}{\DW{x}{1}}{right=of b1}
%       \rf{a3}{b1}
%       \po{b1}{b2}
%       % \rf[out=169,in=11]{b2}{a2}
%       %\rf[out=169,in=11]{b2}{a1}
%       \phevent{p2}{\DR{x}{0}}{below=of a1}
%       \phevent{p1}{\DR{x}{0}}{left=of p2}
%       \mrg{p1}[pos=.3]{a1}
%       \mrg{p2}[right]{a1}      
%       \pox[out=-15,in=-170]{b1}[below]{b2}
%       \pox[out=-15,in=-170]{p1}[below]{p2}
%       \pox{p2}[below]{a3}
%     \end{tikzinline}}
% \end{gather*}

@@

\bookmark{Theorem and Proof Sketch}
\begin{theorem}
  Let $\aPS$ be $L$-stable in $\aCmd$.  Let $\aPS'$ be a $\aCmd$-extension of
  $\aPS$ that is $L$-sequential after $\aPS$.  Let $\aPS''$ be a
  $\aCmd$-extension of $\aPS'$ that is $\rpox$-convex, such that no subset of
  $\Event''$ satisfies these criteria.
  Then either (1) $\aPS''$ is $L$-sequential
  \begin{changed}
    and $L$-simple
  \end{changed}
  after $\aPS$ or (2) there is some $\aCmd$-extension $\aPS'''$ of $\aPS'$
  and some $\aEv\in(\Event''\setminus\Event')$ such that (a) $\aPS'''$ is
  $\aEv$-similar to $\aPS''$, (b) $\aPS'''$ is $L$-sequential
  \begin{changed}
    and $L$-simple
  \end{changed}
  after $\aPS$, and (c) $\bEv\lrace{L}\aEv$, for some
  $\bEv\in(\Event''\setminus\Event)$.
\end{theorem}
The theorem provides an inductive characterization of \emph{Sequential
  Consistency for Local
  Data-Race Freedom (SC-LDRF)}: Any extension of a $L$-stable pomset is either
$L$-sequential, or is $\aEv$-similar to a $L$-sequential extension that
includes a race involving $\aEv$.
\begin{proof}[Proof Sketch]
  \begin{changed}
    We show $L$-sequentiality.  $L$-simplicity then follows from
    \reflem{lem:sequential:simple}.
  \end{changed}

  In order to develop a technique to find $\aPS'''$ from $\aPS''$, we analyze
  pomset order in generation-minimal top-level pomsets.  First, we note that
  $\lt_*$ (the transitive reduction $\lt$) can be decomposed into three
  disjoint relations.  Let ${\rppo}=({\lt_*}\cap{\rpox})$ denote
  \emph{preserved} program order, as required by sequential composition and
  conditional.  The other two relations are cross-thread subsets of
  $({\lt_*}\setminus{\rpox})$: $\rrfe$ (reads-from-external) orders writes
  before reads, satisfying \ref{par-le-rf1} and \ref{par-le-rf2}; $\rcae$
  (coherence-after-external) orders read and write accesses before writes,
  satisfying \ref{pom-rf-block}. (Within a thread, \ref{seq-le-delays-rf} and
  \ref{seq-le-rf-rf} induce order that is included in ${\rppo}$.)

@@

>>>>

% \section{Local Data Race Freedom and Sequential Consistency}
\section{Proof Sketch: LDRF-SC for \PwTmcaTITLE{}}
\label{sec:sc}

%\todo{Remove this section?}

\begin{changed}
  In this appendix, we sketch a proof of \drfsc{} for \PwTmca{2}.  We prove an
  \emph{external} result, where the notion of \emph{data-race} is independent
  of the semantics itself.  Since every \PwTmca{2} is also a \PwTmca{1}, the
  result also applies there.  Our result is also \emph{local}.  Using
  \citeauthor{Dolan:2018:BDR:3192366.3192421}'s
  [\citeyear{Dolan:2018:BDR:3192366.3192421}] notion of \emph{Local Data Race
    Freedom (LDRF)}.

@@

\subsection{Definitions}
\begin{changed}
  \paragraph{Program Order}
  Let $\sempomca{2}{}$ be defined by applying the construction of
  \reflem{lem:po} to $\semmca{2}{}$.  We consider only \emph{complete}
  pomsets.  For these, we derive program order on compound events as follows.
  By \reflem{lem:compound:phantom:exists}, if there is a compound event
  $\aEv$, then there is a phantom event $\cEv\in\fmrginv{\aEv}$ such that
  $\labelingForm(\cEv)$ is a tautology.  If there is exactly one tautology,
  we identify $\aEv$ with $\cEv$ in program order.  If there is more than one
  tautology, \reflem{lem:sequential:simple}, below, shows that it suffices to
  pick an arbitrary one---we identify $\aEv$ with the $\cEv\in\fmrginv{\aEv}$
  that is minimal in program order.
\end{changed}
For example, consider \jmm{} causality test case 2, with an added write to $z$:
\begin{gather*}
  \tag{$\ddagger\ddagger$}\label{eq6}
  \begin{gathered}
    \PR{x}{r}\SEMI
    \PW{z}{1}\SEMI
    \PR{x}{s}\SEMI
    \IF{r{=}s}\THEN \PW{y}{1}\FI
    \PAR
    \PW{x}{y}
    \\
    \hbox{\begin{tikzinline}[node distance=1em and 1.5em]
        \phevent{p1}{\DR{x}{1}}{}
        \event{z}{\DW{z}{1}}{right=of p1}
        \phevent{p2}{\DR{x}{0}}{right=of z}
        \event{a3}{\DW{y}{1}}{right=of p2}
        \event{b1}{\DR{y}{1}}{right=3em of a3}
        \event{b2}{\DW{x}{1}}{right=of b1}
        \rf{a3}{b1}
        \po{b1}{b2}
        \pox[out=-15,in=-170]{p1}{z}
        \pox[out=-15,in=-170]{z}{p2}
        \pox[out=-15,in=-170]{p2}{a3}
        \pox[out=-15,in=-170]{b1}{b2}
        \event{a1}{\DR{x}{1}}{above=of z}
        \mrg{p1}[pos=.3]{a1}
        \mrg{p2}[right]{a1}      
        \rf{b2}{a1}
        \pox{a1}{z}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}
% For TC2, we have: 
% \begin{gather*}
%   \PR{x}{r}\SEMI
%   \PR{x}{s}\SEMI
%   \IF{r{=}s}\THEN \PW{y}{1}\FI
%   \PAR
%   \PW{x}{y}
%   \\
%   \hbox{\begin{tikzinline}[node distance=1.5em]
%       \event{a1}{\DR{x}{1}}{}
%       % \event{a2}{\DR{x}{1}}{right=of a1}
%       \event{a3}{\DW{y}{1}}{right=of a1}
%       % \po{a1}{a3}
%       % \po[out=-20,in=-160]{a1}{a3}
%       \event{b1}{\DR{y}{1}}{right=3em of a3}
%       \event{b2}{\DW{x}{1}}{right=of b1}
%       \rf{a3}{b1}
%       \po{b1}{b2}
%       % \rf[out=169,in=11]{b2}{a2}
%       \rf[out=169,in=11]{b2}{a1}
%       \phevent{p2}{\DR{x}{1}}{below=of a1}
%       \phevent{p1}{\DR{x}{1}}{left=of p2}
%       \mrg{p1}[pos=.3]{a1}
%       \mrg{p2}[right]{a1}      
%       \pox[out=-15,in=-170]{b1}[below]{b2}
%       \pox[out=-15,in=-170]{p1}[below]{p2}
%       \pox{p2}[below]{a3}
%     \end{tikzinline}}
%   \\
%   \hbox{\begin{tikzinline}[node distance=1.5em]
%       \event{a1}{\DR{x}{0}}{}
%       % \event{a2}{\DR{x}{1}}{right=of a1}
%       \event{a3}{\DW{y}{1}}{right=of a1}
%       % \po{a1}{a3}
%       % \po[out=-20,in=-160]{a1}{a3}
%       \event{b1}{\DR{y}{1}}{right=3em of a3}
%       \event{b2}{\DW{x}{1}}{right=of b1}
%       \rf{a3}{b1}
%       \po{b1}{b2}
%       % \rf[out=169,in=11]{b2}{a2}
%       %\rf[out=169,in=11]{b2}{a1}
%       \phevent{p2}{\DR{x}{0}}{below=of a1}
%       \phevent{p1}{\DR{x}{0}}{left=of p2}
%       \mrg{p1}[pos=.3]{a1}
%       \mrg{p2}[right]{a1}      
%       \pox[out=-15,in=-170]{b1}[below]{b2}
%       \pox[out=-15,in=-170]{p1}[below]{p2}
%       \pox{p2}[below]{a3}
%     \end{tikzinline}}
% \end{gather*}

@@

\subsection{Theorem and Proof Sketch}
\begin{theorem}
  Let $\aPS$ be $L$-stable in $\aCmd$.  Let $\aPS'$ be a $\aCmd$-extension of
  $\aPS$ that is $L$-sequential after $\aPS$.  Let $\aPS''$ be a
  $\aCmd$-extension of $\aPS'$ that is $\rpox$-convex, such that no subset of
  $\Event''$ satisfies these criteria.
  Then either (1) $\aPS''$ is $L$-sequential
  \begin{changed}
    and $L$-simple
  \end{changed}
  after $\aPS$ or (2) there is some $\aCmd$-extension $\aPS'''$ of $\aPS'$
  and some $\aEv\in(\Event''\setminus\Event')$ such that (a) $\aPS'''$ is
  $\aEv$-similar to $\aPS''$, (b) $\aPS'''$ is $L$-sequential
  \begin{changed}
    and $L$-simple
  \end{changed}
  after $\aPS$, and (c) $\bEv\lrace{L}\aEv$, for some
  $\bEv\in(\Event''\setminus\Event)$.
\end{theorem}
The theorem provides an inductive characterization of \emph{Sequential
  Consistency for Local
  Data-Race Freedom (SC-LDRF)}: Any extension of a $L$-stable pomset is either
$L$-sequential, or is $\aEv$-similar to a $L$-sequential extension that
includes a race involving $\aEv$.
\begin{proof}[Proof Sketch]
  \begin{changed}
    We show $L$-sequentiality.  $L$-simplicity then follows from
    \reflem{lem:sequential:simple}.
  \end{changed}

  In order to develop a technique to find $\aPS'''$ from $\aPS''$, we analyze
  pomset order in generation-minimal top-level pomsets.  First, we note that
  $\lt_*$ (the transitive reduction $\lt$) can be decomposed into three
  disjoint relations.  Let ${\rppo}=({\lt_*}\cap{\rpox})$ denote
  \emph{preserved} program order, as required by sequential composition and
  conditional.  The other two relations are cross-thread subsets of
  $({\lt_*}\setminus{\rpox})$: $\rrfe$ (reads-from-external) orders writes
  before reads, satisfying \ref{par-le-rf1} and \ref{par-le-rf2}; $\rcae$
  (coherence-after-external) orders read and write accesses before writes,
  satisfying \ref{pom-rf-block}. (Within a thread, \ref{seq-le-delays-rf} 
  %and \ref{seq-le-rf-rf}
  induces order that is included in ${\rppo}$.)

@@

\subsection{A Note on Prior Work}

In preparing this paper, we came across the following example, which appears
to invalidate Theorem 4.1 of \cite{DBLP:conf/ppopp/DongolJR19}.
\begin{gather}
  \nonumber
  \PW{x}{1}\SEMI
  \PW[\mREL]{y}{1}\SEMI
  \PR[\mACQ]{x}{r}
  \PAR
  \IF{\PR[\mACQ]{y}{}}\THEN \PW[\mREL]{x}{2}\FI
  \\
  \tag{\P}
  \label{mix1}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a1}{\DW{x}{1}}{}
      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
      \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
      % \raevent{b1}{\DR[\mACQ]{y}{1}}{below=of a1}
      \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
      \sync{a1}{a2}
      \rf[out=20,in=160]{a1}{a3}
      \rf[out=20,in=160]{a2}{b1}
      \wk[out=20,in=160]{a3}{b2}
      \sync{b1}{b2}
      % \node(ai)[left=3em of a1]{};
      % \bgoval[yellow!50]{(ai)}{P}
      % \bgoval[pink!50]{(a1)(a2)(b1)(b2)}{P'\setminus P}
      % \bgoval[green!10]{(a3)}{P'''\setminus P'}
    \end{tikzinline}}
  \\
  \nonumber
  % \label{mix2}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a1}{\DW{x}{1}}{}
      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
      \raevent{a3}{\DR[\mACQ]{x}{2}}{right=of a2}
      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
      \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
      \sync{a1}{a2}
      \rf[out=20,in=160]{a2}{b1}
      \rf[out=-160,in=-20]{b2}{a3}
      \sync{b1}{b2}
    \end{tikzinline}}
\end{gather}
The program is data-race free.  The two executions shown are the only
top-level executions that include $\DWP[\mREL]{x}{2}$.

Theorem 4.1 of \cite{DBLP:conf/ppopp/DongolJR19} is stated by extending
execution sequences.  In the terminology of
\cite{DBLP:conf/ppopp/DongolJR19}, a read is \emph{$L$-weak} if it is
sequentially stale.  Let
\begin{math}
  \rho=\DWP{x}{1}\allowbreak
  \DWP[\mREL]{y}{1}\allowbreak
  \DRP[\mACQ]{y}{1}\allowbreak
  \DWP[\mREL]{x}{2}
\end{math}
be a sequence and
\begin{math}
  \alpha=\DRP[\mACQ]{x}{1}.
\end{math}
$\rho$ is $L$-sequential and $\alpha$ is $L$-weak in $\rho\alpha$.  But there
is no execution of this program that includes a data race, contradicting the
theorem.  The error seems to be in Lemma A.4 of
\cite{DBLP:conf/ppopp/DongolJR19}, which states that if $\alpha$ is $L$-weak
after an $L$-sequential $\rho$, then $\alpha$ must be in a data race.  That
is clearly false here, since $\DRP[\mACQ]{x}{1}$ is stale, but the program is
data race free.

In proving the SC-LDRF result in \cite[\textsection8]{DBLP:journals/pacmpl/JagadeesanJR20}, we noted that our proof
technique is more robust than that of \cite{DBLP:conf/ppopp/DongolJR19},
because it limits the prefixes that must be considered.  In \eqref{mix1}, the
induction hypothesis requires that we add $\DRP[\mACQ]{x}{1}$ before
$\DWP[\mREL]{x}{2}$ since $\DRP[\mACQ]{x}{1}\xwk\DWP[\mREL]{x}{2}$.  In
particular,
\begin{gather*}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a1}{\DW{x}{1}}{}
      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
      % \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
      % \raevent{b1}{\DR[\mACQ]{y}{1}}{below=of a1}
      \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
      \sync{a1}{a2}
      % \rf[out=20,in=160]{a1}{a3}
      \rf{a2}{b1}
      % \wk[out=20,in=160]{a3}{b2}
      \sync{b1}{b2}
      % \node(ai)[left=3em of a1]{};
      % \bgoval[yellow!50]{(ai)}{P}
      % \bgoval[pink!50]{(a1)(a2)(b1)(b2)}{P'\setminus P}
      % \bgoval[green!10]{(a3)}{P'''\setminus P'}
    \end{tikzinline}}
\end{gather*}
is not a downset of \eqref{mix1}, because
$\DRP[\mACQ]{x}{1}\xwk\DWP[\mREL]{x}{2}$.  As noted in \cite[\textsection8]{DBLP:journals/pacmpl/JagadeesanJR20},
this affects the inductive order in which we move across pomsets, but does
not affect the set of pomsets that are considered.  In particular,
\begin{gather*}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a1}{\DW{x}{1}}{}
      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
      % \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
      % \raevent{b1}{\DR[\mACQ]{y}{1}}{below=of a1}
      % \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
      \sync{a1}{a2}
      % \rf[out=20,in=160]{a1}{a3}
      \rf{a2}{b1}
      % \wk[out=-20,in=-160]{a3}{b2}
      % \sync{b1}{b2}
      % \node(ai)[left=3em of a1]{};
      % \bgoval[yellow!50]{(ai)}{P}
      % \bgoval[pink!50]{(a1)(a2)(b1)(b2)}{P'\setminus P}
      % \bgoval[green!10]{(a3)}{P'''\setminus P'}
    \end{tikzinline}}
\end{gather*}
is a downset of \eqref{mix1}.

### 2021-11-01 paper/intro.tex
accepted!

In certain cases it is possible to ban concurrent access using separation
\cite{OHearn:2007:RCL:1235896.1236121}, or to accept inefficient
implementation in order to obtain sequential consistency (SC)
\cite{DBLP:conf/snapl/MarinoMMNS15}.  When these approaches are not
available, however, the humble semicolon becomes shrouded in mystery, covered
in the cloak of something known as a \emph{memory model}.  Every language has
such a model: For each read operation, it determines the set of available
values.  Compilers and runtime systems are allowed to chose any value in the
set.  To allow efficient implementation, the set must not be too small.  To
allow invariant reasoning, the set must not be too large.

For optimized concurrent languages, it is surprising difficult to define a
model that allows common compiler optimizations and hardware reorderings yet
disallows nonsense behaviors that don't arise in practice.  The latter are
commonly known as ``thin air'' behaviors \cite{DBLP:conf/esop/BattyMNPS15}.
There are only a handful of solutions, and all have deficiencies.  These can
be classified by their approach to dependency tracking (from strongest to
weakest):
\begin{itemize}
\item Syntactic dependencies
  \cite{DBLP:conf/pldi/LahavVKHD17,Boehm:2014:OGA:2618128.2618134,DBLP:conf/oopsla/VafeiadisN13,DBLP:journals/corr/abs-1804-04214}.
  These models require inefficient implementation of relaxed access.  This is
  a non-starter for safe languages like Java and Javascript, and may be an
  unacceptable cost for low-level languages like \cXI{}.
\item Semantic dependencies \cite{%
    Manson:2005:JMM:1047659.1040336, DBLP:conf/esop/JagadeesanPR10,
    DBLP:conf/popl/KangHLVD17, DBLP:journals/pacmpl/ChakrabortyV19,
    DBLP:conf/pldi/LeeCPCHLV20, promising-ldrf}. These models compute
  dependencies operationally using alternate worlds, making it impossible to
  understood a single execution in isolation; they also allow executions that
  violate temporal reasoning (see \textsection\ref{sec:related}).
\item No dependencies, as in \cXI{} \cite{DBLP:conf/esop/BattyMNPS15} and
  Javascript \cite{DBLP:journals/pacmpl/WattRP19}.  This allows thin air
  executions.
\end{itemize}

@@

We extend \PwT{} to a full memory model in \textsection\ref{sec:mca}, based
on \PwP{} \cite{DBLP:journals/pacmpl/JagadeesanJR20}.
\textsection\ref{sec:results} summarizes the results for this model.  In
addition to powering such a bespoke model, the dependency relation calculated
by \PwT{} can also be used with off-the-shelf models.  For example, in
\textsection\ref{sec:c11} we show that it can be used as an $\rsdep$ relation
for \cXI, adapting the approach of \MRD{}
\cite{DBLP:conf/esop/PaviottiCPWOB20}.  % This is only the second semantics
% to distill dependencies down to a relation compatible with the existing
% \cpp{} standard, and the first one to be fully compositional.
\textsection\ref{sec:tool} describes a tool for automatic evaluation of
litmus tests in this model.  \cXI{} allows thin-air in order to avoid
overhead in the implementation of relaxed reads.  Safe language like OCaml
\cite{Dolan:2018:BDR:3192366.3192421} have typically made the opposite
choice, accepting a performance penalty in order to avoid thin-air.  Just as
\PwT{} can be used to strengthen \cXI{}, it could also be used to weaken
these models, allowing optimal lowering for relaxed reads while banning
thin-air.

>>>>

In certain cases it is possible to ban concurrent access using separation
\cite{OHearn:2007:RCL:1235896.1236121}, or to accept inefficient
implementation in order to obtain sequential consistency (SC)
\cite{DBLP:conf/snapl/MarinoMMNS15}.  When these approaches are not
available, however, the humble semicolon becomes shrouded in mystery, covered
in the cloak of something known as a \emph{memory model}.  Every language has
such a model: For each read operation, it determines the set of available
values.  Compilers and runtime systems are allowed to choose any value in the
set.  To allow efficient implementation, the set must not be too small.  To
allow invariant reasoning, the set must not be too large.

For optimized concurrent languages, it is surprising difficult to define a
model that allows common compiler optimizations and hardware reorderings yet
disallows nonsense behaviors that don't arise in practice.  The latter are
commonly known as ``thin-air'' behaviors \cite{DBLP:conf/esop/BattyMNPS15}.
There are only a handful of solutions, and all have deficiencies.  These can
be classified by their approach to dependency tracking (from strongest to
weakest):
\begin{itemize}
\item Syntactic dependencies
  \cite{DBLP:conf/pldi/LahavVKHD17,Boehm:2014:OGA:2618128.2618134,DBLP:conf/oopsla/VafeiadisN13,DBLP:journals/corr/abs-1804-04214}.
  These models require inefficient implementation of relaxed access.  This is
  a non-starter for safe languages like Java and Javascript, and may be an
  unacceptable cost for low-level languages like \cXI{}.
\item Semantic dependencies \cite{%
    Manson:2005:JMM:1047659.1040336, DBLP:conf/esop/JagadeesanPR10,
    DBLP:conf/popl/KangHLVD17, DBLP:journals/pacmpl/ChakrabortyV19,
    DBLP:conf/pldi/LeeCPCHLV20, promising-ldrf}. These models compute
  dependencies operationally using alternate worlds, making it impossible to
  understood a single execution in isolation; they also allow executions that
  violate temporal reasoning (see \textsection\ref{sec:related}).
\item No dependencies, as in \cXI{} \cite{DBLP:conf/esop/BattyMNPS15} and
  Javascript \cite{DBLP:journals/pacmpl/WattRP19}.  This allows thin-air
  executions.
\end{itemize}

@@

We extend \PwT{} to a full memory model in \textsection\ref{sec:mca}, based
on \PwP{} \cite{DBLP:journals/pacmpl/JagadeesanJR20}.
\textsection\ref{sec:results} summarizes the results for this model.  In
addition to powering such a bespoke model, the dependency relation calculated
by \PwT{} can also be used with off-the-shelf models.  For example, in
\textsection\ref{sec:c11} we show that it can be used as an $\rsdep$ relation
for \cXI, adapting the approach of \MRD{}
\cite{DBLP:conf/esop/PaviottiCPWOB20}.  % This is only the second semantics
% to distill dependencies down to a relation compatible with the existing
% \cpp{} standard, and the first one to be fully compositional.
\textsection\ref{sec:tool} describes a tool for automatic evaluation of
litmus tests in this model.  \cXI{} allows thin-air in order to avoid
overhead in the implementation of relaxed reads.  Safe languages like OCaml
\cite{Dolan:2018:BDR:3192366.3192421} have typically made the opposite
choice, accepting a performance penalty in order to avoid thin-air.  Just as
\PwT{} can be used to strengthen \cXI{}, it could also be used to weaken
these models, allowing optimal lowering for relaxed reads while banning
thin-air.
